---
title: "Getting started with boosterpak"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with boosterpak}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`boosterpak` adds a small intent layer above `pak` and `renv`. A project declares
the packs it wants in `boosters.toml`; `boosterpak::sync()` resolves those packs
and installs missing packages into the active project-local `renv` library.

## Initialize

```{r}
boosterpak::init(renv = "yes", rprofile = "yes")
```

This creates:

- `boosters.toml`
- `boosters/packs/`
- `air.toml`, when enabled by settings
- a `.Rprofile` helper-source line, when requested

In non-interactive sessions, use explicit `renv` and `rprofile` arguments. The
default `"ask"` mode never silently changes `.Rprofile` when user confirmation
would be required.

## Add a Pack

```{r}
boosterpak::add_pack("example")
```

The `example` pack installs `cli`. `add_pack()` preserves unrelated comments and
sections in `boosters.toml` while updating `[packs].declared`.

## Apply Project Intent

```{r}
boosterpak::sync()
```

Apply mode is additive in v0.1. It installs missing declared packages and, when
`auto_snapshot = true`, snapshots with `renv`.

## Restore Exact Versions

```{r}
boosterpak::sync(mode = "restore")
```

Restore mode requires both `boosters.toml` and `renv.lock`, calls
`renv::restore()`, validates the manifest, and warns if direct declared packages
are absent from the lockfile.

## Inspect

```{r}
boosterpak::status()
boosterpak::list_packs()
```

`status()` reports whether the manifest exists and validates, what packs and
packages resolve, whether project-local `renv` appears active, whether
`renv.lock` exists, and whether `.Rprofile` contains the helper hook.
