---
title: "Getting started with boosterpak"
format: html
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Getting started with boosterpak}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

`boosterpak` adds a small intent layer above `pak` and `renv`. A project declares
the packs it wants in `boosters.toml`; `boosterpak::sync()` resolves those packs
and installs missing packages into the active project-local `renv` library.
Packs can also carry exact copied helper files, so a useful project setup can be
reused without separating "packages" from customized `boosters/fn_*.R` files.

## Typical workflow

```{mermaid}
flowchart TD
  A[Start or clone project] --> B[Initialize boosterpak]
  B --> C[Add reusable intent]
  C --> D[Add packs]
  C --> E[Add helper functions]
  D --> F[Sync project]
  E --> F
  F --> G[Packages installed in renv]
  F --> H[Helper files copied into boosters]
  G --> I[Edit code and helper files]
  H --> I
  I --> J[Save a reusable pack]
  J --> K[Pack TOML plus function sidecar]
  K --> L[Promote to user scope]
  L --> M[Reuse in another project]
  M --> D
```

In practice, initialize once, add pack and function intent over time, run
`sync()` whenever the project should match that intent, then save and promote a
pack when the setup is worth reusing.

## Initialize

```{r}
#| echo: true
#| eval: false
boosterpak::init(renv = "yes", rprofile = "yes")
```

This creates:

- `boosters.toml`
- `boosters/packs/`
- `air.toml`, when enabled by settings
- a `.Rprofile` helper-source line, when requested

In non-interactive sessions, use explicit `renv` and `rprofile` arguments. The
default `"ask"` mode never silently changes `.Rprofile` when user confirmation
would be required.

## Add a Pack

```{r}
#| echo: true
#| eval: false
boosterpak::add_pack("example")
```

The `example` pack installs `cli`. `add_pack()` preserves unrelated comments and
sections in `boosters.toml` while updating `[packs].declared`.
If the pack carries helper functions, `add_pack()` also copies the bundled
`fn_*.R` files into `boosters/` and refuses to overwrite existing function files
unless `overwrite_functions = TRUE`.

Packs can mix ordinary CRAN package names with package-specific source specs.
Declare every package in `packages`, then use `[sources]` only for packages that
should install from another source:

```toml
name = "plotting"
description = "Plotting packages from CRAN and GitHub."
packages = ["ggplot2", "patchwork", "ggtext"]

[sources]
"ggtext" = "wilkelab/ggtext"
```

Here, `ggplot2` and `patchwork` install by package name, while `ggtext` uses the
GitHub source spec.

## Add a Helper Function

```{r}
#| echo: true
#| eval: false
boosterpak::add_function("ni")
```

Helper functions are materialized as `boosters/fn_*.R` files and tracked in
`[functions].installed`. They are ordinary project files, so you can edit them
after materialization. `sync()` restores missing installed helper files, but it
does not overwrite local edits.

## Apply Project Intent

```{r}
#| echo: true
#| eval: false
boosterpak::sync()
```

Apply mode is additive. It installs missing declared packages, materializes
missing bundled or installed helper functions, and, when `auto_snapshot = true`,
snapshots with `renv`.

## Capture and Reuse

```{r}
#| echo: true
#| eval: false
boosterpak::save_pack("project_baseline")
boosterpak::promote_pack("project_baseline")
```

`save_pack()` captures the resolved package set and, by default, helper functions
listed in `[functions].installed`. Use `functions = "all"` to capture every
`boosters/fn_*.R` file, `functions = "none"` for a package-only pack, or a
character vector for specific function names. `promote_pack()` and
`demote_pack()` copy both the pack TOML and any function sidecar directory.

## Restore Exact Versions

```{r}
#| echo: true
#| eval: false
boosterpak::sync(mode = "restore")
```

Restore mode requires both `boosters.toml` and `renv.lock`, calls
`renv::restore()`, validates the manifest, and warns if direct declared packages
are absent from the lockfile.

## Inspect

```{r}
#| echo: true
#| eval: false
boosterpak::status()
boosterpak::list_packs()
```

`status()` reports whether the manifest exists and validates, what packs and
packages resolve, whether project-local `renv` appears active, whether
`renv.lock` exists, and whether `.Rprofile` contains the helper hook.
