---
title: "Restoring a project on a new machine"
format: html
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Restoring a project on a new machine}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette covers the clone path: a boosterpak project that already works on
one machine, freshly cloned from a git remote onto another. The goal is to get
from `git clone` to a working session without re-running project setup.

The short version:

```{r}
#| echo: true
#| eval: false
# In R, from the cloned project directory:
renv::restore()

# Restart R, then:
boosterpak::sync(mode = "restore")
boosterpak::status()
```

Do not re-run `init()`. The project is already initialized; `sync()` is the
restore verb.

## What travels with the repo

A boosterpak project commits its intent and its lockfile, not its installed
environment. A clone brings:

- `boosters.toml`, the project control file.
- `boosters/packs/`, the project-local pack sidecars.
- `boosters/fn_*.R`, materialized helper functions, including any local edits.
- `boosters/attach.R`, the managed startup attachment file.
- `renv.lock`, the exact package versions.
- `.Rprofile` with the renv activation line and the boosterpak startup hook.
- `renv/activate.R`, the renv bootstrap script.

A clone does not bring:

- The renv project library. Packages must be reinstalled.
- The machine-wide user pack directory. Packs promoted with `promote_pack()`
  live outside the repo, but the project does not need them to restore:
  `add_pack()` already materialized every declared pack into
  `boosters/packs/`.

## 1. Install R and open the project

Install R on the new machine, ideally the version recorded in the `"R"` field
of `renv.lock`. Then start R in the cloned project directory. The committed
`.Rprofile` sources `renv/activate.R`, which bootstraps renv into an empty
project library automatically.

The first startup may print errors from the boosterpak hook: `.Rprofile`
sources `boosters/attach.R`, whose `library()` calls refer to packages that are
not installed yet. These are harmless and disappear after restore.

## 2. Restore the library from the lockfile

`boosterpak` itself is not installed on this machine yet, so nothing
boosterpak-shaped can run first. Restore through renv, which installs
everything in the lockfile, including `boosterpak`. `init()` snapshots
`renv`, `pak`, and `boosterpak` into the lockfile precisely so that a clone can
bootstrap this way.

```{r}
#| echo: true
#| eval: false
renv::restore()
```

## 3. Reconcile the boosterpak layer

Restart R so the freshly restored packages and the startup hook load cleanly,
then run restore mode:

```{r}
#| echo: true
#| eval: false
boosterpak::sync(mode = "restore")
```

Restore mode requires both `boosters.toml` and `renv.lock`, calls
`renv::restore()`, validates the project config, and warns if direct declared
packages are absent from the lockfile. It is lockfile-exact: it does not
hydrate from local libraries, does not install anything beyond the lockfile,
and does not rewrite `boosters/attach.R` or helper files. Helper functions and
`attach.R` travel through git, so restore mode has nothing to materialize;
local edits to `boosters/fn_*.R` are preserved because they are ordinary
committed files.

## 4. Verify

```{r}
#| echo: true
#| eval: false
boosterpak::status()
boosterpak::check_functions()
```

`status()` reports config validity, resolved packs and packages, missing direct
packages, attach state, function drift, renv state, lockfile presence, and the
`.Rprofile` hook. `check_functions()` compares materialized helper files
against their pack sources.

## What does not re-run in a clone

- `init()` side effects. The project already has `boosters.toml`, the
  `.Rprofile` hook, and renv infrastructure; there is nothing to initialize.
- Pack `on_add` hooks. Hooks fire when a pack is newly declared by
  `add_pack()`. In a clone, every pack is already declared in `boosters.toml`,
  so `sync()` restores the environment without re-running project-scaffolding
  side effects such as folder creation.

## Troubleshooting: boosterpak absent from the lockfile

Projects initialized with boosterpak 0.5 may not have `boosterpak` in
`renv.lock`, so `renv::restore()` leaves it uninstalled. Bootstrap it once
without relying on `boosterpak` being loadable:

```{r}
#| echo: true
#| eval: false
install.packages("renv")
renv::install(c("pak", "seanthimons/boosterpak"))
renv::snapshot(packages = c("renv", "pak", "boosterpak"), prompt = FALSE)
```

Then restart R and run `boosterpak::sync()`.
